---
title: "R-Ladies: Introduction to Data Cleaning with `dataMaid`"
author: "Molly Olson, MS"
date: "March 19, 2018"
output: 
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
    toc: yes
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<style type="text/css">

div.main-container {
    max-width: 3200px;
}

.tocify-header {
    text-indent: 0;
}

div.tocify {
  max-width: 100%;
}
</style>


  <script language="javascript"> 
    function toggle(num) {
      var ele = document.getElementById("toggleText" + num);
      var text = document.getElementById("displayText" + num);
      if(ele.style.display == "block") {
        ele.style.display = "none";
        text.innerHTML = "show";
      }
      else {
        ele.style.display = "block";
        text.innerHTML = "hide";
      }
   } 
  </script>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(diagram)
require(kableExtra)
require(knitr)
require(data.table)
```

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 150)
```
# What is the `dataMaid` package?

<a id="displayText" href="javascript:toggle(4);">Show 
</a>  
<div id="toggleText4" style="display: none"> 

Short answer : 

* A data cleaning assistant that is able to provide a document to be read and evaluated by a human. 
* A tool to aid in logic/error checks both column and row-wise 

Long answer : read documentation

</div>

# Goals 

<a id="displayText" href="javascript:toggle(5);">Show 
</a>  
<div id="toggleText5" style="display: none">  


* Data screening/cleaning with the `dataMaid` package
* Extend data cleaning checks with `dataMaid`
* Use `validate` to inspect row-wise errors

</div>


# Examples of errors in data and data cleaning checks

<a id="displayText" href="javascript:toggle(6);">Show 
</a>  
<div id="toggleText6" style="display: none">  

* Incorrect class
* Duplicates
* Capitalization consistency (new York vs New York)
* Unlikely value (BMI = 0.1, age = 201)
* White spaces
* Unrecognized missingness indicators
* Amount of missingness
* Unique observations / categories with low count
* Inaccurate data (death date before birth date)

</div>


# Workflow with `dataMaid`

<a id="displayText" href="javascript:toggle(7);">Show 
</a>  
<div id="toggleText7" style="display: none"> 

```{r, echo = FALSE}
names <- c("Data frame", "Summarize/ \n Visualize", "Check/Inspect", "Validate", "Fix", "Done")

M <- matrix(nrow = 6, ncol = 6, byrow = TRUE, data = c(
#d s  c  v  f   d
0, 0, 0, 0, 0, 0, #d
"", 0, 0, "Errors", "", 0, #s
0, "", 0, 0, 0, 0, #c
0, 0, "Correct", 0, 0, 0, #v
0, 0, "Errors", 0, 0, 0, #f
0, 0, 0, "Correct", 0, 0 #d
))
#

pp <- plotmat(M, pos = c(3,2,1), curve = 0, name = names,
lwd = 1, box.lwd = 2, cex.txt = 0.7,
box.type = "square", box.prop = 0.5, arr.type = "simple", arr.length = 0.55,
arr.pos = 0.5, shadow.size = 0, prefix = "",
main = "")
#

```

Summarize: What information does the variable have?  
Visualize: What does the distribution of the variable look like?  
Check: What potential problems are there with the variable?  
Fix: Fix problems.  
Validate: Did we actually fix the problem and/or check row-wise errors.  

</div>

# Summarize/Visualize


<a id="displayText" href="javascript:toggle(9);">Show 
</a>  
<div id="toggleText9" style="display: none"> 

```{r, results = "asis", echo = FALSE}
static_help <- function(pkg, topic, out, links = tools::findHTMLlinks()) {
  pkgRdDB = tools:::fetchRdDB(file.path(find.package(pkg), 'help', pkg))
  force(links)
  tools::Rd2HTML(pkgRdDB[[topic]], out, package = pkg,
                 Links = links, no_links = is.null(links))
}

tmp <- tempfile()

static_help("dataMaid","makeDataReport",tmp)

out <- readLines(tmp)
headfoot <- grep("body", out)
cat(out[(headfoot[1] + 1):(headfoot[2] - 1)], sep = "\n")
```

</div>

<a id="displayText" href="javascript:toggle(8);">Show 
</a>  
<div id="toggleText8" style="display: none"> 

```{r, echo = TRUE, message = FALSE}
require(dataMaid)
head(bigPresidentData)
bigPresidentData <- as.data.table(bigPresidentData)
```

```{r, eval = FALSE}
makeDataReport(bigPresidentData)
```


**Default report:**  

* Identifies miscoded missing values  
* Identifies prefixed and suffixed (the following appear with the prefix or suffix *x*)  
* Checks whitespaces  
* Identifies levels with < 6 observations  
* Identifies case issues  
* Identifies misclassified numeric or integer variables  
* Identifies outliers  
* Provides a summary table of variable class, number of unique observations, missingness and any problems  
* "Column-wise" checks


</div>

## Using `dataMaid` interactively

<a id="displayText" href="javascript:toggle(12);">Check
</a>  
<div id="toggleText12" style="display: none"> 

```{r, results = "asis"}
allCheckFunctions()
```

```{r}
## syntax

## numeric class
check(bigPresidentData$presidencyYears
      , numericChecks = c("identifyMissing","identifyOutliers"))
check(bigPresidentData$presidencyYears
      , checks = setChecks(numeric = c("identifyMissing")))
check(bigPresidentData$presidencyYears
      , checks = setChecks())

## factor class
check(bigPresidentData$ethnicity
      , factorChecks = c("identifyCaseIssues","identifyLoners"))
check(bigPresidentData$ethnicity
      , checks = setChecks(factor = c("identifyLoners")))
check(bigPresidentData$ethnicity
      , checks = setChecks())

## both
check(bigPresidentData[,.(ethnicity, presidencyYears)]
      , checks = setChecks(numeric = "identifyOutliers"
                         , factor  = "identifyLoners"))
```


```{r, echo = T}
identifyWhitespace(bigPresidentData$firstName)
identifyWhitespace(bigPresidentData$lastName)

check(bigPresidentData[,.(firstName, lastName)]
      , checks = setChecks(character = "identifyWhitespace"))
```

</div>


<a id="displayText" href="javascript:toggle(10);">Visualize 
</a>  
<div id="toggleText10" style="display: none"> 

```{r, results = "asis"}
allVisualFunctions()
```

```{r}
visualize(bigPresidentData$ageAtInauguration, vnam = "Age at inauguration")

basicVisual(bigPresidentData$party, vnam = "Party")
standardVisual(bigPresidentData$party, vnam = "Party") + theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


</div>


<a id="displayText" href="javascript:toggle(11);">Summarize
</a>  
<div id="toggleText11" style="display: none"> 

```{r, results = "asis"}
allSummaryFunctions()
```

```{r, echo = T}
summarize(bigPresidentData$lastName)
# summarize(bigPresidentData)

lapply(bigPresidentData, variableType) %>% head
```

</div>

## Extending `dataMaid`

<a id="displayText" href="javascript:toggle(14);">Show
</a>  
<div id="toggleText14" style="display: none"> 

Customize your own checking, visual, or summary functions!

**Templates:**
```{r, echo = T, eval = F}
mySummaryFunction <- summary(v, ...){
  val <- [ result of whatever summary you want to do ]
  res <- [ properly escaped version of val ]
  summaryResult(list( feature = "[Feature name]"
                    , result  = res
                    , value   = val
                ))
}
```

```{r, echo = T, eval = F}
isSNN <- function(v, nMax = NULL, ...){
  out <- list(problem = FALSE
            , message = ""
            , problemValues = NULL)
  if (class(v) %in% c("character","factor","labelled")){
    if(any(grep("\\d{3}-\\d{3}-\\d{4}",v))) {
      out$problem <- TRUE
      out$message <- "Warning: may contain SSNs"
      out$problemValues <- "Will not show"
    }
  }
  out
}
```

**Examples:**

*Basic example:*
```{r, echo = T}
refCat <- function(v, ...) {
  out <- list(factor = FALSE
            , reference = ""
            , problemValues = NULL)
  if(class(v) %in% c("factor")) {
    out$factor <- TRUE
    out$reference <- levels(v)[1]
    out$problemValues <- "Not applicable"
  }
  out
}

refCat <- summaryFunction(refCat
                        , description = "Identifies reference level"
                        , classes = c("factor"))

check(bigPresidentData$sex, factorChecks = "refCat")
```

*More advanced example:*
```{r, echo = T}
identifyNonStartCase <- function(v, nMax = 10, ...){
  v <- unique(na.omit(v)) ## omit NA values and keep only unique values
  vSplit <- strsplit(v, split = " ") ## split around blank spaces
  vSplitAllLower <- sapply(vSplit, tolower) ## make all lowercase
  helper <- function(x){ ## helper function to make first letter capital
    capFirstLetters <- toupper(substring(x, 1, 1))
    x <- paste(capFirstLetters, substring(x, 2), sep = "")
    x
  }
  vSplitStartCase <- sapply(vSplit, helper) ## first letter capital version of v
  vStartCase <- sapply(vSplitStartCase, function(x) paste(x, collapse = " "))
  
  ## find where v and vStartCase differ
  problemPlaces <- v != vStartCase
  
  if(any(problemPlaces)){
    problemValues <- v[problemPlaces]
  } else {
    problemValues <- NULL
  }
  
  problem <- any(problemPlaces)
  
  problemStatus <- list(problem = problem
                      , problemValues = problemValues)
  problemMessage <- "The following variables were not in start case:"
  outMessage <- messageGenerator(problemStatus, problemMessage, nMax)
  
  checkResult(list(problem = problem
                 , message = outMessage
                 , problemValues = problemValues))
}

identifyNonStartCase <- checkFunction(identifyNonStartCase
                                    , description = "Identifies entries that are not written in start case"
                                    , classes = c("character", "factor"))

check(bigPresidentData$stateOfBirth, checks = setChecks(character = "identifyNonStartCase"))
```

```{r, results = "asis"}
allCheckFunctions()
```

## Customizing document



```{r, echo = FALSE}
#Copy data. This is the data we will make changes to.
bpd <- bigPresidentData
#We fix the following mistake:
#Aragorn Arathornson is included in the dataset.
bpd <- bpd[!(bpd$firstName == "Aragorn" & bpd$lastName == "Arathornson"
),]
#We fix the following mistake:
#Trump has "." listed as his first name (firstName).
bpd[bpd$lastName == "Trump", "firstName"] <- "Donald"
#We fix the following mistake:
#Obama's presidency duration is listed as infinite (presidencyYears).
bpd[bpd$lastName == "Obama", "presidencyYears"] <-
 floor((bpd$presidencyEndDate[bpd$lastName == "Obama"] -
 bpd$presidencyBeginDate[bpd$lastName == "Obama"])/365.25)
#We fix the following mistake:
#Trump's state of birth (New York) was spelled with a lower case "y" (stateOfBirth).
bpd[bpd$lastName == "Trump", "stateOfBirth"] <- "New York"
#We fix the following mistake:
#Truman's last name is prefixed with whitespace (lastName).
bpd[bpd$lastName == " Truman", "lastName"] <- "Truman"
#We fix the following mistake:
#ageAtInauguration is coded as a character variable.
bpd$ageAtInauguration <- as.numeric(bpd$ageAtInauguration)
#We fix the following mistake:
#The dates marking the beginning and end of the presidency
#(presidencyBeginDate, presidencyEndDate) have been switched for Theodore and
#Franklin Roosevelt.
TRbegin <- bpd$presidencyBeginDate[bpd$firstName == "Franklin" &
                                     bpd$lastName == "Roosevelt"]
FRbegin <- bpd$presidencyBeginDate[bpd$firstName == "Theodore" &
                                     bpd$lastName == "Roosevelt"]
TRend <- bpd$presidencyEndDate[bpd$firstName == "Franklin" &
                                 bpd$lastName == "Roosevelt"]

FRend <- bpd$presidencyEndDate[bpd$firstName == "Theodore" & bpd$lastName
 == "Roosevelt"]
bpd$presidencyBeginDate[bpd$firstName == "Franklin" & bpd$lastName == "Ro
osevelt"] <- FRbegin
bpd$presidencyBeginDate[bpd$firstName == "Theodore" & bpd$lastName == "Ro
osevelt"] <- TRbegin
bpd$presidencyEndDate[bpd$firstName == "Franklin" & bpd$lastName == "Roos
evelt"] <- FRend
bpd$presidencyEndDate[bpd$firstName == "Theodore" & bpd$lastName == "Roos
evelt"] <- TRend
#We fix the following mistake:
#Jimmy Carter's end of presidency date (presidencyEndDate) has been changed such that
#his presidency overlaps with the subsequent president, Ronald Reagan.
bpd$presidencyEndDate[bpd$firstName == "Jimmy" & 
                        bpd$lastName == "Carter"] <- as.Date("1975-01-14")
#We fix the following mistake:
#James Garfield's state of birth (stateOfBirth) has been changed from Ohio to Indiana
#(state of birth of Jim Davis, the creator of the cartoon "Garfield").
bpd$stateOfBirth[bpd$firstName == "James" & 
                   bpd$lastName == "Garfield"] <- "Ohio"
#We fix the following mistake:
#Calvin Goolidge has had his first name changed to "Hobbes" (firstName).
bpd$firstName[bpd$firstName == "Hobbes"] <- "Calvin"
##We fix the following mistake:
#Chester Arthur has had his presidency duration (presidencyYears) changed from 3 to 5 years
bpd$presidencyYears[bpd$firstName == "Chester" & 
                      bpd$lastName == "Arthur"] <- 3
#We fix the following mistake:
#Eisenhower appears twice in the dataset, one time with the first name "Dwight"
#and one time with the first name "Dwight D".
#Note: We delete the observation with the extra "D" as the other presidents do
#not have their middle names included
bpd <- bpd[bpd$firstName != "Dwight D",]
#Save a new copy of the data
saveRDS(bpd, file = "bigPresidentData_cleaned.rds")
```

</div>

# Validate

<a id="displayText" href="javascript:toggle(15);">Show 
</a>  
<div id="toggleText15" style="display: none"> 
```{r, message = F}
require(validate)
```

This mostly serves as a logic check (ex. death date is later than birth date) and row-wise checks. 

First, create a `validator` object:
```{r}
validator1 <- validator(
  ageAtDeath := floor((dateOfDeath - birthday)/365.25)
  , `Adult president` = ageAtInauguration >= 18
  , `Alive at inauguration` = ageAtDeath >= ageAtInauguration
  , `Positive first name` = firstName*2 > firstName
  , `Death by assassination` = 
             if (dateOfDeath == presidencyEndDate) 
               assassinationAttempt == 1
  , `Begin date` = difftime(presidencyEndDate, as.Date("176-08-04")) > 0
)
```

```{r}
confront_messy <- confront(bigPresidentData, validator1)
summary(confront_messy) %>% kable

errors(confront_messy)

bpd_clean <- readRDS("bigPresidentData_cleaned.rds")
confront_clean <- confront(bpd_clean, validator1)
summary(confront_clean) %>% kable
```

```{r}
orderVal <- validator(rank(presidencyBeginDate) == rank(orderOfPresidency))

orderCon_messy <- confront(
  na.omit(bigPresidentData[,.(presidencyBeginDate, orderOfPresidency)])
  , orderVal
)
summary(orderCon_messy) %>% kable

orderCon_clean <- confront(
  na.omit(bpd_clean[,.(presidencyBeginDate, orderOfPresidency)])
  , orderVal
)
summary(orderCon_clean) %>% kable
```

Overview of confrontation results:
```{r}
summary(confront_messy)
```
Compute percentage pass/fail/NA:
```{r}
aggregate(confront_messy)
```
Sort results by problem prevalence:
```{r}
sort(confront_messy)
```
For each observation and each check: `TRUE/FALSE/NA`
```{r}
values(confront_messy)
```
Visual overview of check results:
```{r}
barplot(confront_messy)
```
What errors were there:
```{r}
errors(confront_messy)
```
What warnings were there:
```{r}
warnings(confront_messy)
```

</div>

# Make Codebook
<a id="displayText" href="javascript:toggle(16);">Show 
</a>  
<div id="toggleText16" style="display: none"> 

```{r, eval = FALSE}
makeCodebook(bpd_clean)
```

</div>

# Pros and Cons

<a id="displayText" href="javascript:toggle(25);">Show 
</a>  
<div id="toggleText25" style="display: none"> 

## Pros

* Provides a solution and workflow for data cleaning

* Fairly fast even with millions of data

* Can help (prevent) coding errors (`rms` really has trouble with low observations, can find these before hitting errors)

* Helps with collaborations (especially consultations!)
  + Variables in a dataset can usually only be understood in the proper context of their origin 
  + Requires a collaborative effort between an expert in the field and a statistician 
  + Readable to both parties

* Easy for statistician to document what was or was not done

## Cons

* Hard to parse through with lots of data (variables)

* Cannot use non-class dependent variable constraints



</div>

# Resources
Ekstrom, C.T., Peterson, A.H. (2018, February). Cleaning Up the Data Cleaning Process. Short course at the Conference on Statistical Practice, Portland, OR.

Ekstrom, C.T., Peterson, A.H. (2018, February). Cleaning Up the Data Cleaning Process. Short course at the Conference on Statistical Practice, Portland, OR [*Course slides*] .
